---
title: "Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"
author: "Akira Hirao"
date: "`r Sys.Date()`"

output:
  md_document:
    variant: markdown_github
    toc: true
  pdf_document:
    latex_engine: lualatex
    mainfont: "IPAexMincho"
    toc: true
---

# Set ennvironment

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(ThermoSSM)

library(tidyverse)
library(forecast)
library(cowplot)

# initialize
rm(list=ls(all=TRUE))
```

# Loading example data

```{r message = FALSE, warning = FALSE, echo = TRUE}

# loading example data 1: daily mean sea surface temperature time series off southern Ibaraki Prefecture, Japan
data(ibaraki_sst)
head(ibaraki_sst)
```
# Checking example data

```{r message = FALSE, warning = FALSE, echo = TRUE}
class(ibaraki_sst) # zoo object
frequency(ibaraki_sst)   # frequency（1）
start(ibaraki_sst)       # start time（"1982-01-01"）
end(ibaraki_sst)         # end time（"2025-12-31"）
```

# Plotting daily time series of sea surface temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}
daily_ibaraki_sst_plot <- forecast::autoplot(ibaraki_sst) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Daily mean sea-surface temperature off southern Ibaraki Prefecture")

ggsave("Daily_SST_ibaraki_plot.png",
       plot=daily_ibaraki_sst_plot,
       width=8,height=6)

plot(daily_ibaraki_sst_plot)

```

# Converting from daily zoo object to monthly ts object

```{r message = FALSE, warning = FALSE, echo = TRUE}
monthly_ibaraki_sst <- zoo_daily2ts_monthly(ibaraki_sst)
```

# Check of ts object

```{r message = FALSE, warning = FALSE, echo = TRUE}
class(monthly_ibaraki_sst) # ts object
frequency(monthly_ibaraki_sst)   # frequency（12）
start(monthly_ibaraki_sst)       # start time（1982, 1）
end(monthly_ibaraki_sst)         # end time（"2025, 12）
cycle(monthly_ibaraki_sst) %>% head() # 各観測の月番号（1～12）
time(monthly_ibaraki_sst)  %>% head() # 小数年（1982.000, 1982.083...）
window(monthly_ibaraki_sst, start = c(2001, 1), end = c(2001, 12))  # 期間抽出

```

# Plotting monthly time series of sea surface temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}

monthly_ibaraki_sst_plot <- forecast::autoplot(monthly_ibaraki_sst) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Montly mean sea-surface temperature off southern Ibaraki Prefecture")

ggsave("monthly_SST_ibaraki_plot.png",
       plot=monthly_ibaraki_sst_plot,
       width=8,height=6)

plot(monthly_ibaraki_sst_plot)
```

# Plotting time series of temperature deviation

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
monthly_ibaraki_sst_dev <- ThermoSSM::generate_temp_dev(monthly_ibaraki_sst)

monthly_ibaraki_sst_dev_plot <- forecast::autoplot(monthly_ibaraki_sst_dev) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Sea-surface temperature anomaliese")
  
ggsave("monthly_SST_dev_ibaraki.png",
       plot=monthly_ibaraki_sst_dev_plot,
       width=8,height=6)
plot(monthly_ibaraki_sst_dev_plot)
```

# Plotting seasonal pattern of monthly temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}
plot_monthly_SST_seasonal_ibaraki <- ThermoSSM::plot_typical_seasonal_cycle(monthly_ibaraki_sst) 
ggsave("monthly_SST_seasonal_ibaraki.png",
       plot=plot_monthly_SST_seasonal_ibaraki,
       width=8,height=6)
plot(plot_monthly_SST_seasonal_ibaraki)
```


# Applying a Linear Gaussian State-Space Model

```{r message = FALSE, warning = FALSE, echo = TRUE}
res <- lgssm(monthly_ibaraki_sst)
fit    <- res[[1]] # fit model
smooth <- res[[2]] # smoothing

ci <- confint(smooth, level = 0.95) # 95% confident interval
```

# Simple Model Diagnostics

```{r message = FALSE, warning = FALSE, echo = TRUE}

##  model convergence
# OK when return of 0
fit$optim.out$convergence


## normality of residuals
# 標準化残差
std_obs_resid <- rstandard(smooth, type = "recursive")

# forecastパッケージのcheckredisuals関数で残差のチェック
# Ljung–Box検定: P > 0.05で残差に有意な自己相関なしと判断
checkresiduals(std_obs_resid)

# 図示された残差（上：残差系列；左下：残差コレログラム；右下：残差のヒストグラム）をみて異常に突出した残差がないかなどを確認


## normality of residuals
# P > 0.05で正規分布と有意に異なっていないと判断
shapiro.test(std_obs_resid)

```

# Parameter Estimation of the Model

```{r message = FALSE, warning = FALSE, echo = TRUE}

# 推定パラメーター(過程誤差、観測誤差など)
pars_comp <- ThermoSSM::extract_pars(res)
pars_comp


# 平滑化推定量
alpha_hat <- smooth$alphahat
head(alpha_hat)


#　水準成分の平滑化推定量
level_ts <- ThermoSSM::extract_level(res)

#　ドリフト成分の平滑化推定量
drift_ts <- ThermoSSM::extract_drift(res)

# 年あたりの平均的な昇温率
mean_drift_year <- mean(drift_ts) * 12
print(mean_drift_year)

# 1990年代における年あたりの平均昇温率
drift_1990s_per_year <- window(drift_ts,
                               start=c(1990,1),
                               end=c(1999,12)
                               ) %>%  mean()*12
print(drift_1990s_per_year)

# 2010年代における年あたりの平均昇温率
drift_2010s_per_year <- window(drift_ts,
                               start=c(2010,1),
                               end=c(2019,12)
                               ) %>%  mean()*12
print(drift_2010s_per_year)

```

# Plotting estimates of the level, drift, seasonal, and auto-regression components

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_comp <- ThermoSSM::plot_level_trend_season_ar(res)
plot_four_comp <- cowplot::plot_grid(plt_comp[[1]],
                   plt_comp[[2]],
                   plt_comp[[3]],
                   plt_comp[[4]],
                   ncol=1) 

ggsave("plot_level_trend_season_ar.png",
       plot=plot_four_comp,
       width=8,height=8)
plot_four_comp
```

# Plotting level component with confidence interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_ci <- ThermoSSM::plot_level_ci(res,
                              ci_range = 0.95)

ggsave("level_plot.png",
       width=6, height=4,
       plot = plt_level_ci)

plot(plt_level_ci)
```

# Plotting drift component with confidence interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_drift_ci <- ThermoSSM::plot_drift_ci(res,
                              ci_range = 0.95)

ggsave("drift_plot.png",
       width=6, height=4,
       plot = plt_drift_ci)

plot(plt_drift_ci)
```

# Handling original temperature data

```{r message = FALSE, warning = FALSE, echo = TRUE}
##========================
## Example 1
# Monthly temperature data
original_data <- data.frame(
  Year=c(2000,2000,2000,2000,2000,2000,2000,2000,2000,2000,2000,2000),
  Month=c(1,2,3,4,5,6,7,8,9,10,11,12),
  Temp=   c(5.2, 6.1, 9.3, 13.5, 17.8, 21.0,24.3, 25.1, 22.0, 17.1, 11.2,7.0))

head(original_data)

# Convert ts object
original_ts <- ts(
  matrix(original_data$Temp),
  start = c(original_data$Year[1],original_data$Month[1]),
  frequency = 12
)

# Important!!: Exe colnames() to add "Temp" name to ts object for the model 
colnames(original_ts) <- c("Temp")

frequency(original_ts)
start(original_ts)
end(original_ts)
time(original_ts)
original_ts


##========================
## Example 2
# loading and checking raw csv data
input_csv <- "./data/test_data.csv"
raw_csv <- read_csv(input_csv)
head(raw_csv)

# Load original csv file and then convert to monthly ts object 
test_ts <- ThermoSSM::monthly_csv2ts(input_csv)
head(test_ts)

frequency(test_ts)
start(test_ts)
end(test_ts)
time(test_ts)


```
