---
title: "Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"
author: "Akira Hirao"
date: "`r Sys.Date()`"

output:
  md_document:
    variant: markdown_github
    toc: true
  pdf_document:
    latex_engine: lualatex
    mainfont: "IPAexMincho"
    toc: true
---

# Set ennvironment

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(ThermoSSM)

library(tidyverse)
library(forecast)
library(cowplot)

# initialize
rm(list=ls(all=TRUE))
```

# Loading test data

```{r message = FALSE, warning = FALSE, echo = TRUE}

# loading monthly mean air-temperature at the summit of Mt. Fuji in Japan
data(fuji_temp)
head(fuji_temp)
```
# Check of ts object

```{r message = FALSE, warning = FALSE, echo = TRUE}

frequency(fuji_temp)   # frequency（12）
start(fuji_temp)       # start time（1982, 1）
end(fuji_temp)         # end time（2025, 10）
cycle(fuji_temp) %>% head() # 各観測の月番号（1～12）
time(fuji_temp)  %>% head() # 小数年（1982.000, 1982.083...）
window(fuji_temp, start = c(2001, 1), end = c(2001, 12))  # 期間抽出
```

# Plotting time series of temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}

fuji_temp_plot <- forecast::autoplot(fuji_temp) +
  labs(y = expression(Temperature~(degree*C)), x = "Time")

plot(fuji_temp_plot)
#ggsave("Fuji_temp_plot.png",
#       width=8,height=4)
```

# Plotting time series of temperature deviation

```{r message = FALSE, warning = FALSE, echo = TRUE}

plot_temp_dev_obj <- plot_temp_dev(fuji_temp) 
#ggsave("Fuji_temp_dev_plot.png",
#       width=8,height=4)
```

# Plotting seasonal pattern of monthly temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}
plot_typical_seasonal_cycle(fuji_temp) 
#ggsave("Fuji_ave_monthly_temp_plot.png",
#       width=8,height=6)
```


# Applying a Linear Gaussian State-Space Model

```{r message = FALSE, warning = FALSE, echo = TRUE}
res <- lgssm(fuji_temp)
fit    <- res[[1]] # fit model
smooth <- res[[2]] # smoothing

ci <- confint(smooth, level = 0.95) # 95% confident interval
```

# Simple Model Diagnostics

```{r message = FALSE, warning = FALSE, echo = TRUE}

##  model convergence
# OK when return of 0
fit$optim.out$convergence


## normality of residuals
# 標準化残差
std_obs_resid <- rstandard(smooth, type = "recursive")

# forecastパッケージのcheckredisuals関数で残差のチェック
# Ljung–Box検定: P > 0.05で残差に有意な自己相関なしと判断
checkresiduals(std_obs_resid)

# 図示された残差（上：残差系列；左下：残差コレログラム；右下：残差のヒストグラム）をみて異常に突出した残差がないかなどを確認


## normality of residuals
# P > 0.05で正規分布と有意に異なっていないと判断
shapiro.test(std_obs_resid)

```

# Parameter Estimation of the Model

```{r message = FALSE, warning = FALSE, echo = TRUE}

# 推定パラメーター(過程誤差、観測誤差など)
pars_comp <- extract_pars(res)
pars_comp


# 平滑化推定量
alpha_hat <- smooth$alphahat
head(alpha_hat)

level <- alpha_hat[,"level"]
drift <- alpha_hat[,"slope"]

level_ts <- ts(level, start = start(fuji_temp), frequency = 12)
drift_ts <- ts(drift, start = start(fuji_temp), frequency = 12)

# 年あたりの平均的な昇温率
mean_drift_year <- mean(drift_ts) * 12
print(mean_drift_year)

# 1940年代における年あたりの平均昇温率
drift_1940s_per_year <- window(drift_ts,
                               start=c(1940,1),
                               end=c(1949,12)
                               ) %>%  mean()*12
print(drift_1940s_per_year)

# 2010年代における年あたりの平均昇温率
drift_2010s_per_year <- window(drift_ts,
                               start=c(2010,1),
                               end=c(2019,12)
                               ) %>%  mean()*12
print(drift_2010s_per_year)

```

# Plotting estimates of the level, drift, seasonal, and auto-regression components

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_comp <- ThermoSSM::plot_level_trend_season_ar(res)
plot_four_comp <- cowplot::plot_grid(plt_comp[[1]],
                   plt_comp[[2]],
                   plt_comp[[3]],
                   plt_comp[[4]],
                   ncol=1) 

ggsave("plot_level_trend_season_ar.png",
       plot=plot_four_comp,
       width=8,height=8)
plot_four_comp
```

# Plotting level component with confidence interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_ci <- plot_level_ci(res,
                              ci_range = 0.95)

ggsave("level_plot.png",
       width=6, height=4,
       plot = plt_level_ci)

plot(plt_level_ci)
```

# Plotting drift component with confidence interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_drift_ci <- plot_drift_ci(res,
                              ci_range = 0.95)

ggsave("drift_plot.png",
       width=6, height=4,
       plot = plt_drift_ci)

plot(plt_drift_ci)
```


