---
title: "Quick tutorial for linear Gaussian state-space analysis of temperature time series with R package tempssm"
author: "Akira Hirao"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
#    toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm
vignette: >
  %\VignetteIndexEntry{"Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

R package **tempssm** provides tools for state-space analysis of temperature time series.
It provides tools for fitting linear Gaussian state-space models and conducting
inference via Kalman filtering and smoothing, implemented using the
KFAS R package (Helske, 2017).   

## Key features

- Designed for temperature time series with arbitrary seasonal frequencies;
  currently validated primarily on monthly data
- Estimates latent states using linear Gaussian state-space models
  combined with Kalman filtering and smoothing
- Models temperature dynamics as a sum of interpretable latent components,
  including a long-term trend, seasonal cycle, autoregressive structure,
  and optional exogenous effects
- Allows users to specify an arbitrary order of the autoregressive component
  (default: AR(1))
- Implements time-series cross-validation for model evaluation

## Input Data Format

Input data for **tempssm** must be supplied as an R `ts` object,
which represents a regularly spaced time series
(see `?stats::ts` or
<https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html>).

To support data preparation, the package includes utility functions
that convert external observational data into `ts` objects suitable
for model fitting (see Appendix).

## To Install
```r
if(!require("devtools"))
	install.packages("devtools")
devtools::install_github("akihirao/tempssm")
```

## Set Ennvironment for Practices

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(tempssm)
library(forecast)

library(purrr) 
library(tibble)
library(readr)
library(dplyr)
library(ggplot2)

library(future)
library(parallelly)

has_future <- requireNamespace("future", quietly = TRUE)

if (has_future) {
  if (future::nbrOfWorkers() == 1) {

    # minimqnized setting for vignette
    workers <- min(2, future::availableCores())

    if (future::supportsMulticore()) {
      future::plan(future::multicore, workers = workers)
    } else {
      future::plan(future::multisession, workers = workers)
    }

  }
} else {
  message("future not available; running sequentially")
}

```

## Practice I: Applying a Linear Gaussian State-Space Model
## to a Univariate Temperature Time Series

### Objective

The objective of this practice is to demonstrate the basic application
of a linear Gaussian state-space model to a univariate temperature time
series. This example serves as an introduction to the modeling framework
and highlights the role of autoregressive dynamics without the inclusion
of exogenous variables.

### Loading the Dataset

A sample sea surface temperature (SST) dataset is included in the package.

- **Dataset**: Monthly sea surface temperature (SST) off Niigata, Japan  
- **Unit**: Degrees Celsius  
- **Period**: February 2002 to December 2023  
``

This dataset is derived from observations archived at the
Japan Oceanographic Data Center (JODC),
Hydrographic and Oceanographic Department, Japan Coast Guard.
The original daily SST data were obtained from  
<https://www.jodc.go.jp/jodcweb/JDOSS/index.html>  
and subsequently aggregated into monthly means.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(niigata_sst) # load a ts object of SST off Niigata
head(niigata_sst)
summary(niigata_sst)
```
The dataset includes two missing observations, which are retained and
handled explicitly within the state-space modeling framework.

### Plotting the Monthly SST Time Series

We begin by visualizing the monthly SST time series to examine its
overall structure, including apparent trends, seasonal variability,
and missing observations.
```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_niigata_sst <- forecast::autoplot(niigata_sst) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Monthly SST off Niigata, Japan") +
  theme_classic()

plot(plt_niigata_sst)
```

### Applying a Linear Gaussian State-Space Model

Here, we apply linear Gaussian state-space models to the univariate SST
time series and examine the effect of the autoregressive (AR) order on
model behavior.   
Specifically, three models are fitted with the order of the
autoregressive component varying from 1 to 3, while all other model
components, including the explicit seasonal cycle, are kept the same.   

By comparing models with different AR orders, we assess how short- and
longer-term temporal dependencies are represented within the
state-space framework.   
The model summaries provide parameter estimates and diagnostics that
can be used to evaluate the adequacy of different autoregressive orders.
Model comparison and selection are discussed in the following section.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# model with first-order autoregressive component
res <- tempssm(niigata_sst) # first order of auto-regressive model (ar_order=1: default)
summary(res)

# model with second-order autoregressive component
res_ar2 <- tempssm(niigata_sst,ar_order=2) 
summary(res_ar2)

# model with third-order autoregressive component
res_ar3 <- tempssm(niigata_sst,ar_order=3) 
summary(res_ar3)

```

### Model selection based on AIC

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Extract AIC
AIC_ar1 <- extract_AIC(res)
AIC_ar2 <- extract_AIC(res_ar2)
AIC_ar3 <- extract_AIC(res_ar3)

AIC_table_res <- tibble(model=c("AR1","AR2","AR3"),
                        AIC = c(AIC_ar1,AIC_ar2,AIC_ar3)) %>% 
  mutate(delta_AIC = AIC - min(AIC))

AIC_table_res %>% knitr::kable()
```

AR1 model is better than the other models.

### Plotting Level, Drift, Seasonal, and Auto-Regressive Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot each of components at once
plot(res)
```

### Simple Model Diagnostics

```{r message = FALSE, warning = FALSE, echo = TRUE}
resid_test_output <- tempssm::wrapper_checkresiduals(res)
print(resid_test_output)
```
Autocorrelation of residuals was not significant by Ljung-Box test (P > 0.05).


### Estimated Parameters and Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res$kfs$alphahat
head(alpha_hat)

#　Smoothing estimate of level component
level_ts <- tempssm::extract_level_ts(res)

#　Smoothing estimate of drift component
drift_ts <- tempssm::extract_drift_ts(res)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts) 
print(mean_drift_year)

# Average drift rate per year from Jan 2006 to Dec 2010
ave_drift_2006_2010 <- window(drift_ts,
                              start=c(2006,1),
                               end=c(2010,12)
                               ) %>%  mean()
print(ave_drift_2006_2010)

# Average drift rate per year from Jan 2011 to Dec 2015
ave_drift_2011_2015 <- window(drift_ts,
                              start=c(2011,1),
                               end=c(2015,12)
                               ) %>%  mean()
print(ave_drift_2011_2015)

# Average drift rate per year from Jan 2016 to Dec 2020
ave_drift_2016_2020 <- window(drift_ts,
                              start=c(2016,1),
                               end=c(2020,12)
                               ) %>%  mean()
print(ave_drift_2016_2020)

```

### Plotting Level and Drift Components with 95% Confidence Interval

We visualize the estimated long-term evolution of temperature levels
and their rates of change (drift) by extracting the corresponding latent
components from the state-space model.
Seasonal variability and autoregressive dependence are separated out,
allowing the underlying trend behavior to be examined more clearly.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_ci <- plot(res,
                     components = "level",
                     ci = TRUE,
                     ci_level = 0.95
                     ) +
  theme_classic()

plt_drift_ci <- plot(res,
                     components = "drift",
                     ci = TRUE,
                     ci_level = 0.95
                     ) +
  geom_hline(yintercept = 0, lty="dotted") +
  theme_classic()


plt_level_drift_ci <- plt_level_ci + plt_drift_ci + patchwork::plot_layout(ncol=1)

plot(plt_level_drift_ci)
```

The level component shows a persistent upward trend in sea surface
temperature over the study period. The drift component indicates 
a relatively stable positive rate of change, with an estimated 
average annual increase of approximately 0.0525 °C. The shaded 
gray areas represent 95% confidence intervals for the
estimated latent states, illustrating the uncertainty associated 
with the inferred long-term trend and its rate of change.


## Practice II: Applying a Linear Gaussian State-Space Model

## to a Temperature Time Series with an Exogenous Variable

### Objective

The objective of this practice is to use a long record of temperature
time series data to investigate and quantify the influence of a
large-scale climate mode on long-term temperature variations.
Specifically, we examine the effect of the North Atlantic Oscillation
(NAO) as an exogenous variable within a state-space modeling framework.

### Loading Dataset 1: Long-Term Air Temperature Record

- **Data**: Monthly air temperature at the Hohenpeissenberg
  Meteorological Observatory, Germany  
- **Location**: 47.801°N, 11.011°E  
- **Unit**: Degrees Celsius  
- **Period**: January 1781 to December 2025

This dataset represents one of the longest continuous instrumental
records of air temperature available worldwide, observed at a mountain
meteorological station. The original data were obtained from the Global
Historical Climatology Network Monthly (GHCNm) version 4, distributed by
the National Centers for Environmental Information (NCEI), NOAA:
<https://www.ncei.noaa.gov/data/global-historical-climatology-network-monthly/v4>.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(hmo_temp) # loading a ts object of air temperature at the HMO station
head(hmo_temp)
```

### Loading Dataset 2: NAO Index as an Exogenous Variable

- **Data**: Monthly North Atlantic Oscillation (NAO) index (Hurrell)  
- **Period**: January 1865 to June 202

Monthly NAO indices derived using related data　sources are available
from the Climate Analysis Section, National Center for Atmospheric
Research (NCAR), Boulder, USA (Hurrell, 2003). The dataset used in this
package was obtained from:
<https://www.ncei.noaa.gov/access/monitoring/nao/>.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(nao) # load a ts object of NAO index
head(nao)
```

### Intersecting the Temperature and NAO Time Series

For state-space modeling with exogenous variables, all input time series
must share a common and aligned time index. In this step, the
temperature and NAO time series are restricted to their overlapping
period by trimming the leading and trailing portions, ensuring that both
datasets cover an identical time span.

The function `ts.intersect()` is used to align the two `ts` objects on a
shared timeline, returning a multivariate time series containing only
the common period.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate an object on a shared timeline
hmo_nao_ts <- ts.intersect(hmo_temp, nao)
colnames(hmo_nao_ts) <- c("Temp", "NAO")

start(hmo_nao_ts)
end(hmo_nao_ts)

# Generate an ts object of HMO air-temperature with the shared timeline
hmo_temp_common <- hmo_nao_ts[,"Temp"]
start(hmo_temp_common)
end(hmo_temp_common)

# Generate an ts object of NAO with the shared timeline
nao_common <- hmo_nao_ts[,"NAO"]

# Execute label_ts_mono() function to label exogenous variable(s)!!
nao_common <- label_ts_mono(nao_common, label="NAO")
start(nao_common)
end(nao_common)

```

### Plotting Time Series of Air Temperature and NAO Index

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_homo_temp <- forecast::autoplot(hmo_temp_common) +
  labs(x = "Time", y = expression(Temperature~(degree*C))) +
  ggtitle("Air temperature at Hohenpeissenberg Meteorological Observatory") +
  theme_classic()

plt_nao <- forecast::autoplot(nao_common) +
  labs(x = "Time", y = "NAO index") +
  ggtitle("NAO index") +
  theme_classic()

cowplot::plot_grid(plt_homo_temp,
                   plt_nao,
                   ncol=1)

```

### Applying a Model Without an Exogenous Variable

We first fit a baseline state-space model that does not include any
exogenous variables.
This model serves as a reference case in which temperature variability
is explained solely by the latent trend, seasonal cycle, and
autoregressive dependence.

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_without <- tempssm(hmo_temp_common) 
summary(res_without)
```

The fitted model converges successfully and includes a first-order
autoregressive [AR(1)] component.
Prior testing of autoregressive orders from AR(1) to AR(3) indicated
that AR(1) provided the best model fit for both the baseline model and
the model including the exogenous PDO index.
For brevity, the detailed results of this comparison are omitted here;
interested users are encouraged to explore alternative AR orders within
their own analyses.

This baseline model provides a useful benchmark for evaluating the
additional explanatory power of the PDO index introduced in the
following section.

### Applying Model With an Exogenous Variable

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_with <- tempssm(temp_data = hmo_temp_common,
                    exo_data = nao_common) 
summary(res_with)
```

The estimated coefficient for the exogenous NAO index is positive
(0.29), and its 95% confidence interval does not include zero \[0.24,
0.34\], indicating a statistically significant relationship between NAO
variability and local air temperature at the Hohenpeissenberg station.

Specifically, the model suggests that a one-unit increase in the NAO
index is associated with an average increase of approximately 0.29 °C in
monthly air temperature, after accounting for the underlying trend,
seasonal cycle, and autoregressive dependence. This result implies that
positive phases of the NAO contribute systematically to warmer
temperature conditions at the study site.

Importantly, this effect is identified in addition to the internal
dynamics of the temperature time series, rather than being an artifact
of an improved representation of the trend or temporal dependence.
Compared with the baseline model without exogenous variables, the
statistical significance of the NAO coefficient indicates that the NAO
acts as an independent large-scale climate driver influencing local
temperature variability.

We next compare the overall goodness of fit between the two models using
the Akaike Information Criterion (AIC).

### Model comparison based on AICs

Model selection criteria such as the AIC provide a quantitative measure
of model adequacy that balances goodness of fit against model
complexity. Lower AIC values indicate a more parsimonious model with
better support from the data.
```{r message = FALSE, warning = FALSE, echo = TRUE}
AIC_without <- extract_AIC(res_without)
AIC_with <- extract_AIC(res_with)

models_AICs <- tibble(
  model = c("Without","With"),
  AIC = c(AIC_without,AIC_with),
  delta_AIC = min(AIC)-AIC
  )

models_AICs %>% knitr::kable()
```

The model including the NAO index as an exogenous variable exhibits a
substantially lower AIC than the baseline model without exogenous
variables, indicating a markedly better overall fit. This result
supports the conclusion that explicitly accounting for NAO variability
improves the statistical description of the temperature time series
beyond what is achieved by internal dynamics alone.

It should be noted that AIC-based comparisons are meaningful only among
models fitted to the same time series over an identical period.
Moreover, in state-space models, AIC differences reflect both regression
effects and changes in the stochastic structure of latent components.
Therefore, AIC should be used as a complementary diagnostic alongside
parameter estimates and their uncertainty, rather than as a sole basis
for inference.

To further assess the robustness and predictive performance of the
models, we employ time-series cross-validation as described bellow.

### Plotting Level and Drift Components with 95% CI 

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_without_ts <- plot(res_without, 
                             components=c("level"),
                             ci=TRUE) +
  labs(title="Model without the NAO index") +
  theme_classic()

plt_drift_without_ts <- plot(res_without, 
                             components=c("drift"),
                             ci=TRUE) + 
  labs(title="Model without the NAO index") +
  theme_classic()

plt_level_drift_without_ts <- plt_level_without_ts + 
  plt_drift_without_ts + 
  patchwork::plot_layout(ncol=1)

plot(plt_level_drift_without_ts)


plt_level_with_ts <- plot(res_with,
                          components=c("level"),
                          ci=TRUE)+ 
  labs(title="Model with the NAO index") +
  theme_classic()

plt_drift_with_ts <- plot(res_with,
                          components=c("drift"),
                          ci=TRUE) + 
  labs(title="Model with the NAO index") + 
  theme_classic()

plt_level_drift_with_ts <- plt_level_with_ts + 
  plt_drift_with_ts + 
  patchwork::plot_layout(ncol=1)

plot(plt_level_drift_with_ts)
```

Gray areas in the above graph show 95% confidence interval.

### Time Series Cross-Validation (tsCV)

Time series cross-validation (tsCV) approach evaluates model performance 
based on out-of-sample prediction errors while respecting the temporal 
ordering of the data, and thus provides an additional, independent 
perspective on model adequacy.

Time-series cross-validation is conducted by repeatedly fitting the
model to an expanding training window and evaluating its predictive
performance on subsequent observations.
Unlike random cross-validation, this procedure avoids information
leakage from the future to the past and is therefore well suited for
temporal data.

By comparing cross-validation metrics for models with and without the
exogenous PDO variable, we assess whether the improvement suggested by
AIC is also reflected in out-of-sample predictive skill.
``

```{r message = FALSE, warning = FALSE, echo = TRUE}
# ts cross-validation of the model without exogenous variables

## Generate a list of training and test datasets with their indices
# Procedure for constructing year-based time-series cross-validation folds:
# 
# First training data: January 1865–December 1950;
# First test data: one year starting from January 1951
# 
# Second training data: January 1875–December 1960;
# Second test data: one year starting from January 1961
# ...
# 
# Eighth training data: January 1875–December 2020;
# Eighth test data: one year starting from January 2021
#
# These folds are automatically generated by the ts_cv_folds() function.
# Generate training and test dataset

folds_without <- ts_cv_folds(
  temp_data = hmo_temp_common,
  exo_data = NULL,
  initial = 1032, # 1032 monthly observations from Jan 1865 to Dec 1950
  horizon = 12, # forecast 12 monthly time-series
  step = 120, # training data is moved by 120 months (10 years) steps
  fixed_window = FALSE,
  allow_partial = FALSE
  )

folds_with <- ts_cv_folds(
  temp_data = hmo_temp_common,
  exo_data = nao_common,
  initial = 1032, # 1032 monthly observations from Jan 1865 to Dec 1950
  horizon = 12, # forecast 12 monthly time-series
  step = 120, # training data is moved by 12 months (10 years) steps
  fixed_window = FALSE,
  allow_partial = FALSE
)

## Performe tsCV for the model without the exogenous variable of NAO
cv_meta_without <- rolling_origin_tsCV(folds_without,
                                       fold_ids=seq(1:8), 
                                       ar_order=1) 
# Hold on a few minutes 

cv_without <- map(cv_meta_without, pluck, "CV", .default = NULL)

cv_without_tidy <- cv_without %>%
  map(~ as.data.frame(.x) %>%tibble::rownames_to_column(var = "set")) %>%
  purrr::list_rbind(names_to = "cv_id") %>%
  mutate(Model="Without") %>%
  dplyr::relocate(Model,cv_id, set)  # 識別列を先頭に

print(cv_without_tidy)



## Performe tsCV for the model with the exogenous variable of NAO
cv_meta_with <- rolling_origin_tsCV(folds_with,
                                    fold_ids=seq(1:8),
                                    ar_order=1)
# Hold on a few minutes 

cv_with <- map(cv_meta_with, pluck, "CV", .default = NULL)


cv_with_tidy <- cv_with %>%
  map(~ as.data.frame(.x) %>%tibble::rownames_to_column(var = "set")) %>%
  purrr::list_rbind(names_to = "cv_id") %>%
  mutate(Model="With") %>%
  dplyr::relocate(Model,cv_id, set)  # Reorder columns to place identifier variables first

print(cv_with_tidy)

cv_tidy <- bind_rows(cv_without_tidy,cv_with_tidy)
cv_tidy$Model <- factor(cv_tidy$Model,
                        levels=c("Without","With"))

plot_MAE <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MAE)) +
  geom_boxplot()

plot_MAPE <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MAPE)) +
  geom_boxplot()

plot_MASE_naive <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MASE_naive)) +
  geom_boxplot()

plot_MASE_snaive <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MASE_snaive)) +
  geom_boxplot()

cowplot::plot_grid(plot_MAE,plot_MAPE,plot_MASE_naive,plot_MASE_snaive,nrow=1)
```

Analysis of tsCV shows that the model with the exogenous variable is better than 
the model without one. In this tutorial, the number of tsCV iterations has been 
set to eigth to reduce execution time. Please ensure you perform a sufficient 
number of iterations during actual validation.

Taken together, the results consistently support the inclusion of the
NAO index as an exogenous variable in the state-space model.
The NAO coefficient is statistically significant, indicating a clear
local effect on temperature variability, while the improvement in AIC
demonstrates enhanced overall model fit. Furthermore, time-series 
cross-validation confirms that this improvement translates into 
superior out-of-sample predictive performance.

These complementary lines of evidence provide a robust and
multifaceted justification for adopting the exogenous-variable model.

## Appendix: Utility Functions

### Utility function: `monthly_temp_csv2ts()`

The function `monthly_temp_csv2ts()` converts monthly temperature data
stored in a CSV file into an R `ts` object.
It is intended to facilitate the ingestion of externally prepared
time-series data into tempssm by enforcing a simple and consistent
data format.

#### Example

For monthly temperature data, prepare a CSV file with a header row.
By default, the column names must be `Year`, `Month`, and `Temp`,
where `Temp` represents the observed temperature value.

```text
Year,Month,Temp
2010,8,13.6
2010,9,6.8
2010,10,NA
2010,11,-1.4
...
```

* Use NA for missing temperature values, and always keep the corresponding Year and Month entries.
* The CSV file must be comma-separated and UTF-8 encoded.

The following example uses a sample CSV file included in the package.

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")
example_temp <- monthly_temp_csv2ts(path)
head(example_temp)
```

### Utility function: `monthly_temp_df2ts()`

The function `monthly_temp_df2ts()` converts a data frame containing
monthly temperature data into an R `ts` object. It is designed to support
workflows where temperature data are first imported or prepared as a data
frame before being used for time-series analysis.   

The input data frame is expected to contain at least two columns:
a date column (`Date`) and a temperature column (`Temp`).

#### Example

```{r message = FALSE, warning = FALSE}
# Load example CSV file included in the package
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")

# Create a data frame with Date and Temp columns
example_temp_df <- readr::read_csv(path) %>%
  dplyr::mutate(Date = as.Date(paste0(Year, "-", Month, "-01"))) %>%
  dplyr::select(Date, Temp)

# Convert the data frame to a ts object
example_temp_ts <- monthly_temp_df2ts(example_temp_df)

head(example_temp_ts)
```

### Utility function: `sst_jma2ts()`

The function `sst_jma2ts()` downloads publicly available daily mean
sea-surface temperature (SST) data for Japanese coastal waters
provided by the Japan Meteorological Agency (JMA).
It aggregates the daily values into monthly means and returns
the resulting time series as an object of class `ts`.

#### Example

The following example demonstrates how to download SST data for
the southern coastal waters of Ibaraki Prefecture, Japan.

The argument `sea_area_id` specifies the numeric identifier of a sea area
defined by JMA. For example, `138` corresponds to the coastal waters
off southern Ibaraki. A list of available sea area IDs and their
corresponding regions is provided by JMA at:

<https://www.data.jma.go.jp/kaiyou/data/db/kaikyo/series/engan/eg_areano.html>

```{r message = FALSE, warning = FALSE, echo = TRUE}
sst_138_ts <- tempssm::sst_jma2ts(sea_area_id = 138)
head(sst_138_ts)
```

### Utility function: `monthly_anomaly()`

The function `monthly_anomaly()` computes monthly anomalies from a
time series provided as an R `ts` object.
Monthly anomalies are calculated by subtracting the corresponding
long-term monthly mean from each observation, thereby removing the
seasonal cycle while preserving interannual variability.

The reference period used to compute the monthly climatology can be
specified via a function argument; by default, the climatology is
computed over the entire available time series (see `?monthly_anomaly`).

This transformation is useful for exploratory analysis and modeling
applications that focus on departures from typical seasonal conditions.

#### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
data(niigata_sst)
niigata_sst_anomaly <- monthly_anomaly(niigata_sst)
plot(niigata_sst_anomaly, ylab=expression(Anomaly~(degree*C))) 
```

### Utility function: `mean_seasonal_cycle()`

The function `mean_seasonal_cycle()` computes the climatological mean
seasonal cycle from a `ts` object by averaging each seasonal value
across years. It is primarily intended for exploratory analysis and
visualization of the seasonal structure in temperature time series.

#### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(niigata_sst)
monthly_seasonal_cycle_niigata_sst <- mean_seasonal_cycle(niigata_sst) 
summary(monthly_seasonal_cycle_niigata_sst)

plt_monthly_seasonal_cycle_niigata_sst <- ggplot(
  data = monthly_seasonal_cycle_niigata_sst,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of temperature",
    x = "Month",
    y = expression(Temperature~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  ) +
  theme_classic()


plot(plt_monthly_seasonal_cycle_niigata_sst)
```

These utility functions are provided to support data preparation and　exploratory analysis and are not required for the core modeling functionality of tempssm.

## References

The statistical modeling framework implemented in `tempssm`
is based on the methodology described in Baba et al. (2024),
with implementation details adapted from the accompanying
supplementary materials and code repository.   

Baba, S., Ishii, H., and Yoshiyama, T. (2024).
Estimating sea temperature trends using a linear Gaussian state-space model
in Jogashima, Kanagawa, Japan.
*Bulletin of the Japanese Society of Fisheries Oceanography*,
88(3), 190–199.
(In Japanese with an English abstract.)
https://doi.org/10.34423/jsfo.88.3_190

Baba, S. (2024).
Supplementary code and test data for estimating sea temperature trends
using a linear Gaussian state-space model.
GitHub repository:
https://github.com/logics-of-blue/sea-temperature-trend-jogashima

Helske, J. (2017).
KFAS: Exponential Family State Space Models in R.   
*Journal of Statistical Software*, 78(10), 1–39.
https://doi.org/10.18637/jss.v078.i10
