---
title: "Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"
author: "Akira Hirao"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
#    toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm
vignette: >
  %\VignetteIndexEntry{"Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Set Ennvironment

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(ThermoSSM)
library(forecast)
library(tidyverse)
```

# Loading Example Data

```{r message = FALSE, warning = FALSE, echo = TRUE}

# loading example data 1
# Daily mean sea surface temperature time series off southern Ibaraki Prefecture, Japan
data(ibaraki_sst)
head(ibaraki_sst)
```
# Checking Example Data

```{r message = FALSE, warning = FALSE, echo = TRUE}
class(ibaraki_sst) # zoo object
frequency(ibaraki_sst)   # frequency（1）
start(ibaraki_sst)       # start time（"1982-01-01"）
end(ibaraki_sst)         # end time（"2025-12-31"）
```

# Plotting Daily Time Series of Sea Surface Temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}
daily_ibaraki_sst_plot <- forecast::autoplot(ibaraki_sst) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Daily mean sea-surface temperature off southern Ibaraki Prefecture")

plot(daily_ibaraki_sst_plot)
```

# Converting from Daily zoo object to Monthly ts object

```{r message = FALSE, warning = FALSE, echo = TRUE}
monthly_ibaraki_sst <- zoo_daily2ts_monthly(ibaraki_sst)
colnames(monthly_ibaraki_sst) <- "Temp"
```

# Check of ts object

```{r message = FALSE, warning = FALSE, echo = TRUE}
class(monthly_ibaraki_sst) # ts object
frequency(monthly_ibaraki_sst)   # frequency（12）
start(monthly_ibaraki_sst)       # start time（1982, 1）
end(monthly_ibaraki_sst)         # end time（"2025, 12）
cycle(monthly_ibaraki_sst) %>% head() # 各観測の月番号（1～12）
time(monthly_ibaraki_sst)  %>% head() # 小数年（1982.000, 1982.083...）
window(monthly_ibaraki_sst, start = c(2001, 1), end = c(2001, 12))  # 期間抽出

```

# Plotting Monthly Time Series of Sea Surface Temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}

monthly_ibaraki_sst_plot <- forecast::autoplot(monthly_ibaraki_sst) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Montly mean sea-surface temperature off southern Ibaraki Prefecture")

plot(monthly_ibaraki_sst_plot)
```

# Plotting Time Series of Temperature Anomalies

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
monthly_ibaraki_sst_anomaly <- ThermoSSM::monthly_anomaly(monthly_ibaraki_sst)

monthly_ibaraki_sst_anomaly_plot <- forecast::autoplot(monthly_ibaraki_sst_anomaly) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Sea-surface temperature anomaliese")
  
plot(monthly_ibaraki_sst_anomaly_plot)
```

# Plotting Monthly Seasonal Cycle

```{r message = FALSE, warning = FALSE, echo = TRUE}
monthly_seasonal_cycle_SST_ibaraki <- ThermoSSM::mean_seasonal_cycle(monthly_ibaraki_sst) 
summary(monthly_seasonal_cycle_SST_ibaraki)

plt_monthly_seasonal_cycle_SST_ibaraki <- ggplot(
  data = monthly_seasonal_cycle_SST_ibaraki,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of SST",
    x = "Month",
    y = expression(Temperature~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  )


plot(plt_monthly_seasonal_cycle_SST_ibaraki)
```

# Applying a Linear Gaussian State-Space Model

```{r message = FALSE, warning = FALSE, echo = TRUE}
res <- lgssm(monthly_ibaraki_sst)

print(res)
summary(res)
```

# Plotting Level, Drift, Seasonal, and Auto-Regressive Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot each of components at once
plot(res)
```

# Simple Model Diagnostics

```{r message = FALSE, warning = FALSE, echo = TRUE}

## normality of residuals
# 標準化残差
std_obs_resid <- rstandard(res$kfs, type = "recursive")

# forecastパッケージのcheckredisuals関数で残差のチェック
# Ljung–Box検定: P > 0.05で残差に有意な自己相関なしと判断
checkresiduals(std_obs_resid)

# 図示された残差（上：残差系列；左下：残差コレログラム；右下：残差のヒストグラム）をみて異常に突出した残差がないかなどを確認


## normality of residuals
# P > 0.05で正規分布と有意に異なっていないと判断
shapiro.test(std_obs_resid)

```

# Estimated Parameters and Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
#過程誤差、観測誤差などのパラメーター
params <- ThermoSSM::extract_param(res)
params


# 平滑化推定量
alpha_hat <- res$kfs$alphahat
head(alpha_hat)


#　水準成分の平滑化推定量
level_ts <- ThermoSSM::extract_level_ts(res)

#　ドリフト成分の平滑化推定量
drift_ts <- ThermoSSM::extract_drift_ts(res)

# 年あたりの平均的な昇温率
mean_drift_year <- mean(drift_ts) * 12
print(mean_drift_year)

# 1990年代における年あたりの平均昇温率
drift_1990s_per_year <- window(drift_ts,
                               start=c(1990,1),
                               end=c(1999,12)
                               ) %>%  mean()*12
print(drift_1990s_per_year)

# 2010年代における年あたりの平均昇温率
drift_2010s_per_year <- window(drift_ts,
                               start=c(2010,1),
                               end=c(2019,12)
                               ) %>%  mean()*12
print(drift_2010s_per_year)

```

# Plotting Level Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_ci <- plot(res,
                     components = "level",
                     ci = TRUE,
                     ci_level = 0.95
                     )

plot(plt_level_ci)
```

# Plotting Drift Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_drift_ci <- plot(res,
                     components = "drift",
                     ci = TRUE,
                     ci_level = 0.95
                     )

plot(plt_drift_ci)
```

# Appendix: Handling Raw Temperature Data
## This appendix describes how the original temperature time series data are prepared prior to the state-space analysis presented in the main text. We outline the procedures for handling raw observations.

### Example 1

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Monthly temperature data
original_data <- data.frame(
  Year=c(2000,2000,2000,2000,2000,2000,2000,2000,2000,2000,2000,2000),
  Month=c(1,2,3,4,5,6,7,8,9,10,11,12),
  Temp=   c(5.2, 6.1, 9.3, 13.5, 17.8, 21.0,24.3, 25.1, 22.0, 17.1, 11.2,7.0))

head(original_data)

# Convert ts object
original_ts <- ts(
  matrix(original_data$Temp),
  start = c(original_data$Year[1],original_data$Month[1]),
  frequency = 12
)

# Important!!: Exe colnames() to add "Temp" name to ts object for the model 
colnames(original_ts) <- c("Temp")

frequency(original_ts)
start(original_ts)
end(original_ts)
time(original_ts)
original_ts

res_original <- lgssm(original_ts)
summary(res_original)
```

### Example 2

```{r message = FALSE, warning = FALSE, echo = TRUE}
# loading example csv
# It contains a monthly temperature time series measured at Mount Akadake, Hokkaido, Japan
# The data originated from: https://www.biodic.go.jp/moni1000/findings/data/index.html
# example_csv_path <- system.file("extdata", "example_monthly_temp.csv", package = "ThermoSSM")
example_csv_path <- system.file("extdata", "example_monthly_temp.csv", package = "ThermoSSM")
example_temp <- readr::read_csv(example_csv_path)
head(example_temp)

# Converting tidy object to ts object
example_ts <- ts(matrix(example_temp$Temp),
  start = c(example_temp$Year[1],example_temp$Month[1]),
  frequency = 12
)
colnames(example_ts) <- "Temp"
head(example_ts)

frequency(example_ts)
start(example_ts)
end(example_ts)

res_example <- lgssm(example_ts)
summary(res_example)

plot(res_example)
```
