---
title: "Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"
author: "Akira Hirao"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
#    toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm
vignette: >
  %\VignetteIndexEntry{"Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Set Ennvironment

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(ThermoSSM)
library(forecast)

library(purrr)   # list_rbind
library(tibble)
library(readr)
library(dplyr)
library(ggplot2)

library(future)
library(parallelly)

has_future <- requireNamespace("future", quietly = TRUE)

if (has_future) {
  if (future::nbrOfWorkers() == 1) {

    # vignetteでは控えめに
    workers <- min(2, future::availableCores())

    if (future::supportsMulticore()) {
      future::plan(future::multicore, workers = workers)
    } else {
      future::plan(future::multisession, workers = workers)
    }

  }
} else {
  message("future not available; running sequentially")
}

```

# Practice I: Analysis of Monthly Sea Surface Temperature (SST)

## Loading dataset

### Data: Monthly sea surface temperature (SST) off Niigata, Japan
### Unit: Celsius
### Duration: Feb 2002 to December 2023
### This dataset provided by Japan Oceanographic Data Center (JODC), Hydrographic and Oceanographic Department, Japan Coast Guard. Original daily SST data were obtained from
https://www.jodc.go.jp/jodcweb/JDOSS/index.html and aggregated into monthly means.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(niigata_sst) # load a ts object of SST off Niigata

head(niigata_sst)
```

## Plotting Monthly Time Series of Air Temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_niigata_sst <- forecast::autoplot(niigata_sst) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Monthly SST off Niigata, Japan") +
  theme_classic()

plot(plt_niigata_sst)
# The example data includes missing observation values
```
## Plotting Time Series of Temperature Anomalies

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
niigata_sst_anomaly <- ThermoSSM::monthly_anomaly(niigata_sst)

niigata_sst_anomaly_tidy <- tibble(time=time(niigata_sst_anomaly),
                           anomaly = as.numeric(niigata_sst_anomaly)) %>%
  mutate(sign = ifelse(anomaly >= 0, "Positive","Negative"))


plt_niigata_sst_anomaly <- ggplot() +
  geom_col(data = niigata_sst_anomaly_tidy,
           aes(x = time, y = anomaly, fill = sign),
           width = 0.8,alpha=0.2) +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(
    values = c(
      "Positive"  = "firebrick",
      "Negative" = "steelblue"
    )
  ) +
  geom_line(data = niigata_sst_anomaly_tidy,
            aes(x = time, y = anomaly, fill=""), 
            color="black",alpha = 0.5,
            linewidth = 0.5) +
  guides(fill = "none") +
  labs(
    y = expression(Temperature~(degree*C)), 
    x = "Year"
  )  +
  theme_classic()


plot(plt_niigata_sst_anomaly)
```

## Plotting Monthly Seasonal Cycle

```{r message = FALSE, warning = FALSE, echo = TRUE}
monthly_seasonal_cycle_niigata_sst <- ThermoSSM::mean_seasonal_cycle(niigata_sst) 
summary(monthly_seasonal_cycle_niigata_sst)

plt_monthly_seasonal_cycle_niigata_sst <- ggplot(
  data = monthly_seasonal_cycle_niigata_sst,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of temperature",
    x = "Month",
    y = expression(Temperature~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  ) +
  theme_classic()


plot(plt_monthly_seasonal_cycle_niigata_sst)
```

## Applying a Linear Gaussian State-Space Model

```{r message = FALSE, warning = FALSE, echo = TRUE}
# a model with a first-order autoregressive component
res <- lgssm(niigata_sst) # first order of auto-regressive model (ar_order=1: default)
summary(res)

res_ar2 <- lgssm(niigata_sst,ar_order=2) 
summary(res_ar2)

res_ar3 <- lgssm(niigata_sst,ar_order=3) 
summary(res_ar3)

```
## Model selection based on AIC

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Extract AIC
AIC_ar1 <- extract_AIC(res)
AIC_ar2 <- extract_AIC(res_ar2)
AIC_ar3 <- extract_AIC(res_ar3)

AIC_table_res <- tibble(model=c("AR1","AR2","AR3"),
                        AIC = c(AIC_ar1,AIC_ar2,AIC_ar3)) %>% 
  mutate(delta_AIC = AIC - min(AIC))

AIC_table_res %>% knitr::kable()

## AR1 model is better than the other models.
```


## Plotting Level, Drift, Seasonal, and Auto-Regressive Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot each of components at once
plot(res)
```

## Simple Model Diagnostics

```{r message = FALSE, warning = FALSE, echo = TRUE}
resid_test_output <- ThermoSSM::wrapper_checkresiduals(res)

print(resid_test_output)
```

## Estimated Parameters and Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res$kfs$alphahat
head(alpha_hat)


#　Smoothing estimate of level component
level_ts <- ThermoSSM::extract_level_ts(res)

#　Smoothing estimate of drift component
drift_ts <- ThermoSSM::extract_drift_ts(res)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts) 
print(mean_drift_year)

# Average drift rate per year from Jan 2006 to Dec 2010
ave_drift_2006_2010 <- window(drift_ts,
                              start=c(2006,1),
                               end=c(2010,12)
                               ) %>%  mean()
print(ave_drift_2006_2010)

# Average drift rate per year from Jan 2011 to Dec 2015
ave_drift_2011_2015 <- window(drift_ts,
                              start=c(2011,1),
                               end=c(2015,12)
                               ) %>%  mean()
print(ave_drift_2011_2015)

# Average drift rate per year from Jan 2016 to Dec 2020
ave_drift_2016_2020 <- window(drift_ts,
                              start=c(2016,1),
                               end=c(2020,12)
                               ) %>%  mean()
print(ave_drift_2016_2020)

```

## Plotting Level Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_ci <- plot(res,
                     components = "level",
                     ci = TRUE,
                     ci_level = 0.95
                     ) +
  theme_classic()

plot(plt_level_ci)
```

## Plotting Drift Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_drift_ci <- plot(res,
                     components = "drift",
                     ci = TRUE,
                     ci_level = 0.95
                     ) +
  geom_hline(yintercept = 0, lty="dotted") +
  theme_classic()

plot(plt_drift_ci)
```

# Practice II: Analysis of the Air Temperature with an exogenous variable

## Objective: Using the world’s longest temperature time series from a mountainous region—collected continuously since the mid-18th century along the northern edge of the German Alps—to isolate and verify the effect of of the North Atlantic Decadal Oscillation (NAO) as an exogenous covariate on long-term temperature variations.

## Loading Air Temperature at the Mountaton Station

### Data: Monthly air temperature at Hohenpeissenberg Meteorological Observatory, German
### Unit: Celsius
### Duration: Jan 1781 to December 2025
### This dataset is the longest record of air temperature observed in Mountain region across the world. Original dataset provided by Global Historical Climatology Network monthly (GHCNm) ver.4 was obtained from https://www.ncei.noaa.gov/data/global-historical-climatology-network-monthly/v4

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(hmo_temp) # loading a ts object of air temperature at the HMO station

head(hmo_temp)
```

## loading dataset of NAO index as an exogenous valiable

### Data: Monthly North Atlantic Oscillation (NAO) index (Hurrell)
### Duration: Jan 1865 to Jun 2023
### The winter (December thru March) station-based index of the NAO is based on the difference of normalized sea level pressure (SLP) between Lisbon, Portugal and Stykkisholmur/Reykjavik, Iceland since 1864. Positive values of the NAO index are typically associated with stronger-than-average westerlies over the middle latitudes, more intense weather systems over the North Atlantic and wetter/milder weather over western Europe. Monthly, seasonal and annual indices using slightly different data sources for the southern station are also available.
### Source dataset provided by the Climate Analysis Section, NCAR, Boulder, USA, Hurrell (2003) was obtained from https://www.jodc.go.jp/jodcweb/JDOSS/index.html.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(nao) # load a ts object of NAO index

head(nao)
```

## Intersecting the two ts objects of temperature and NAO

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate an object on a shared timeline
hmo_nao_ts <- ts.intersect(hmo_temp, nao)
colnames(hmo_nao_ts) <- c("Temp", "NAO")

start(hmo_nao_ts)
end(hmo_nao_ts)

# Generate an ts object of HMO air-temperature with the shared timeline
hmo_temp_common <- hmo_nao_ts[,"Temp"]
start(hmo_temp_common)
end(hmo_temp_common)

# Generate an ts object of NAO with the shared timeline
nao_common <- hmo_nao_ts[,"NAO"]

# Execute label_ts_mono() function to label exogenous variable(s)!!
nao_common <- label_ts_mono(nao_common, label="NAO")
start(nao_common)
end(nao_common)

```

## Plotting Time Series of air temperature and NAO index

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_homo_temp <- forecast::autoplot(hmo_temp_common) +
  labs(x = "Time", y = expression(Temperature~(degree*C))) +
  ggtitle("Air temperature at Hohenpeissenberg Meteorological Observatory") +
  theme_classic()

plt_nao <- forecast::autoplot(nao_common) +
  labs(x = "Time", y = "NAO index") +
  ggtitle("NAO index") +
  theme_classic()

cowplot::plot_grid(plt_homo_temp,
                   plt_nao,
                   ncol=1)

```

## Applying a Linear Gaussian State-Space Model without exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_without <- lgssm(hmo_temp_common) 
summary(res_without)
AIC_without <- extract_AIC(res_without)
plot(res_without)
```

## Applying a Linear Gaussian State-Space Model with exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_with <- lgssm(temp_data = hmo_temp_common,exo_data = nao_common) 
summary(res_with)
AIC_with <- extract_AIC(res_with)
plot(res_with)

```

## Model comparison based on AICs

```{r message = FALSE, warning = FALSE, echo = TRUE}

models_AICs <- tibble(
  model = c("Without","With"),
  AIC = c(AIC_without,AIC_with),
  delta_AIC = min(AIC)-AIC
  )

models_AICs %>% knitr::kable()

## Model with the exogenous variable of NAO is remarkably better as shown in the below table. 
```

## time series Cross-Validation

### In this tutorial, the number of tsCV iterations has been set to eigth to reduce execution time. Please ensure you perform a sufficient number of iterations during actual validation.
```{r message = FALSE, warning = FALSE, echo = TRUE}
# ts cross-validation of the model without exogenous variables

## Generate a list of training and test datasets with their indices
# Procedure for constructing year-based time-series cross-validation folds:
# 
# First training data: January 1865–December 1950;
# First test data: one year starting from January 1951
# 
# Second training data: January 1875–December 1960;
# Second test data: one year starting from January 1961
# ...
# 
# Eighth training data: January 1875–December 2020;
# 15th test data: one year starting from January 2021
#
# These folds are automatically generated by the ts_cv_folds() function.
# Generate training and test dataset
folds_without <- ts_cv_folds(
  temp_data = hmo_temp_common,
  exo_data = NULL,
  initial = 1032, # 228 monthly observations from Jan 1865 to Dec 1950
  horizon = 12, # forecast 12 monthly time-series
  step = 120, # training data is moved by 120 months (10 years) steps
  fixed_window = FALSE,
  allow_partial = FALSE
  )

folds_with <- ts_cv_folds(
  temp_data = hmo_temp_common,
  exo_data = nao_common,
  initial = 1032, # 228 monthly observations from Jan 1865 to Dec 1950
  horizon = 12, # forecast 12 monthly time-series
  step = 120, # training data is moved by 12 months (10 years) steps
  fixed_window = FALSE,
  allow_partial = FALSE
)

## Performe tsCV for the model without NAO index
cv_meta_without <- rolling_origin_tsCV(folds_without,
                                       fold_ids=seq(1:8), 
                                       ar_order=1) 
# Hold on a few minutes 

cv_without <- map(cv_meta_without, pluck, "CV", .default = NULL)

cv_without_tidy <- cv_without %>%
  map(~ as.data.frame(.x) %>%tibble::rownames_to_column(var = "set")) %>%
  purrr::list_rbind(names_to = "cv_id") %>%
  mutate(Model="Without") %>%
  dplyr::relocate(Model,cv_id, set)  # 識別列を先頭に

print(cv_without_tidy)



## Performe tsCV for the model with NAO index
cv_meta_with <- rolling_origin_tsCV(folds_with,
                                    fold_ids=seq(1:8),
                                    ar_order=1)
# Hold on a few minutes 

cv_with <- map(cv_meta_with, pluck, "CV", .default = NULL)


cv_with_tidy <- cv_with %>%
  map(~ as.data.frame(.x) %>%tibble::rownames_to_column(var = "set")) %>%
  purrr::list_rbind(names_to = "cv_id") %>%
  mutate(Model="With") %>%
  dplyr::relocate(Model,cv_id, set)  # Reorder columns to place identifier variables first

print(cv_with_tidy)

cv_tidy <- bind_rows(cv_without_tidy,cv_with_tidy)
cv_tidy$Model <- factor(cv_tidy$Model,
                        levels=c("Without","With"))

plot_MAE <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MAE)) +
  geom_boxplot()

plot_MAPE <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MAPE)) +
  geom_boxplot()

plot_MASE_naive <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MASE_naive)) +
  geom_boxplot()

plot_MASE_snaive <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MASE_snaive)) +
  geom_boxplot()

cowplot::plot_grid(plot_MAE,plot_MAPE,plot_MASE_naive,plot_MASE_snaive,nrow=1)
```

# Appendix: Utility Function

## Utility function of sst_jma2ts()

### This function downloads publicly available daily mean sea-surface temperature (SST) data
### for Japanese coastal waters provided by the Japan Meteorological Agency (JMA),
### and returns the corresponding monthly mean time series as an object of class ts.

```{r message = FALSE, warning = FALSE, echo = TRUE}
#' @param sea_area_id
#' Numeric sea area ID. The default is NULL.
#' For example, 138 corresponds to the coastal waters off southern Ibaraki, Japan.
#' A list of sea area IDs and their corresponding regions is available at:
#' \url{https://www.data.jma.go.jp/kaiyou/data/db/kaikyo/series/engan/eg_areano.html}
```

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Download SST data for the southern coastal waters of Ibaraki Prefecture, Japan
sst_138_ts <- ThermoSSM::sst_jma2ts(sea_area_id = 138) 
head(sst_138_ts)

```

