---
title: "Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"
author: "Akira Hirao"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
#    toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm
vignette: >
  %\VignetteIndexEntry{"Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Set Ennvironment

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(ThermoSSM)
library(forecast)

library(purrr)   # list_rbind
library(tibble)
library(readr)
library(dplyr)
library(ggplot2)
```

# Practice I: Analysis of Monthly Air Temperature

## Example Data 1: Monthly air temperature at the summit of Mt. Fuji, Japan
### Duration: July 1937 to December 2025
### This dataset consists of publicly available monthly air temperature data released by the Japan 
### Meteorological Agency (JMA) (https://www.jma.go.jp/jma/indexe.html).

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(fuji_temp) 
# A ts object of monthly air temperature at the summit of Mt. Fuji, Japan
head(fuji_temp)
```

## Plotting Monthly Time Series of Air Temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}
plot_temp_fuji <- forecast::autoplot(fuji_temp) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Monthly mean air temperature at the summit of Mt. Fuji, Japan")

plot(plot_temp_fuji)
# The example data includes missing observation values
```
## Plotting Time Series of Temperature Anomalies

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
example_anomaly <- ThermoSSM::monthly_anomaly(fuji_temp)

monthly_example_anomaly_plot <- forecast::autoplot(example_anomaly) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Temperature anomaliese")
  
plot(monthly_example_anomaly_plot)
```

## Plotting Monthly Seasonal Cycle

```{r message = FALSE, warning = FALSE, echo = TRUE}
monthly_seasonal_cycle_fuji <- ThermoSSM::mean_seasonal_cycle(fuji_temp) 
summary(monthly_seasonal_cycle_fuji)

plt_monthly_seasonal_cycle_fuji <- ggplot(
  data = monthly_seasonal_cycle_fuji,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of temperature",
    x = "Month",
    y = expression(Temperature~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  )


plot(plt_monthly_seasonal_cycle_fuji)
```

## Applying a Linear Gaussian State-Space Model

```{r message = FALSE, warning = FALSE, echo = TRUE}
res <- lgssm(fuji_temp) 
summary(res)
```

## Plotting Level, Drift, Seasonal, and Auto-Regressive Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot each of components at once
plot(res)
```

## Simple Model Diagnostics

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Normality of standardized residuals
std_obs_resid <- stats::rstandard(res$kfs, type = "recursive")

# Residual diagnostics
forecast::checkresiduals(std_obs_resid)
# Ljung–Box test: if P > 0.05, there is no evidence of significant autocorrelation in the residuals
# Inspect the diagnostic plots 
# top: residual time series;
# bottom left: residual autocorrelation;
# bottom right: histogram of residuals to check for outliers or other anomalies
```

## Estimated Parameters and Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Parameters including process errors and observation error
params <- ThermoSSM::extract_param(res)
params

# Smoothing estimates
alpha_hat <- res$kfs$alphahat
head(alpha_hat)


#　Smoothing estimate of level component
level_ts <- ThermoSSM::extract_level_ts(res)

#　Smoothing estimate of drift component
drift_ts <- ThermoSSM::extract_drift_ts(res)

# Average drift rate per year
mean_drift_year <- mean(drift_ts) * 12
print(mean_drift_year)

# Average drift rate per year during 1950s
drift_per_year_1950s <- window(drift_ts,
                               start=c(1950,1),
                               end=c(1959,12)
                               ) %>%  mean()*12
print(drift_per_year_1950s)

# Average drift rate per year during 2010s
drift_per_year_2010s <- window(drift_ts,
                               start=c(2010,1),
                               end=c(2019,12)
                               ) %>%  mean()*12
print(drift_per_year_2010s)

```

## Plotting Level Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_ci <- plot(res,
                     components = "level",
                     ci = TRUE,
                     ci_level = 0.95
                     )

plot(plt_level_ci)
```

## Plotting Drift Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_drift_ci <- plot(res,
                     components = "drift",
                     ci = TRUE,
                     ci_level = 0.95
                     )

plot(plt_drift_ci)
```

# Practice II: Analysis of Monthly Sea-Surface Temperature with Exogenous Variables

## Dataset is originated from Baba et al. (2024)

## Baba, S., Ishii, H., and Yoshiyama, T. (2024).  
## Estimating sea temperature trends using a linear Gaussian state-space model in Jogashima, Kanagawa, Japan.
## Bulletin of the Japanese Society of Fisheries Oceanography, 88, 190–199 (in Japanese with English abstruct)
## https://doi.org/10.34423/jsfo.88.3_190
## Supplementary code and test data are available on GitHub repository:  
## https://github.com/logics-of-blue/sea-temperature-trend-jogashima

## loading dataset and converting ts object

```{r message = FALSE, warning = FALSE, echo = TRUE}
baba_data <- readr::read_csv("https://raw.githubusercontent.com/logics-of-blue/sea-temperature-trend-jogashima/refs/heads/main/src/ssm-sample-data.csv")

# time　　　：日付(論文では0.5か月だが、本サンプルコードでは月単位データとした)
# temp　　　：水温
# distance　：黒潮流軸までの距離
# kuroshio_a：黒潮流路がA型のときに1をとるフラグ
head(baba_data)

# 秋冬の黒潮Aフラグの追加
baba_data <- 
  baba_data %>% 
  mutate(winter = as.numeric(lubridate::month(as.Date(time)) %in% c(11, 12, 1, 2, 3)),
         winter_a = winter * kuroshio_a)

# winter　：秋冬に1をとるフラグ
# winter_a：秋冬かつ、黒潮がAであるときに1をとるフラグ
head(baba_data)

# ts型に変換
# 0.5か月単位データの場合はfrequency = 24と指定する
baba_data_ts <- 
  baba_data %>% 
  dplyr::select(-time, -winter) %>%             # 日付列と秋冬フラグを削除
  ts(start = c(1998, 1), frequency = 12) # ts型に変換(1998年1月開始。12か月1周期)
head(baba_data_ts)
```

## Splitting temperature data and exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
# ts object of temperature time series
baba_temp_ts <- baba_data_ts[,"temp"]

# ts object of exogenous variables
baba_exo_ts <- baba_data_ts[,c("distance","kuroshio_a", "winter_a")]

```

## Plotting Time Series of Temperature 

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_baba_temp <- forecast::autoplot(baba_temp_ts) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Monthly mean sea-surface temperature")

plot(plt_baba_temp)
```

## Plotting Time Series of Exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_baba_exo_distance <- forecast::autoplot(baba_exo_ts[,"distance"]) +
  labs(x = "Time", y = "Distance") +
  ggtitle("Distance")

plt_baba_exo_kuroshio_a <- forecast::autoplot(baba_exo_ts[,"kuroshio_a"]) +
  labs(x = "Time", y = "Category") +
  ggtitle("kuroshio_a")

plt_baba_exo_winter_a <- forecast::autoplot(baba_exo_ts[,"winter_a"]) +
  labs(x = "Time",y = "Category") +
  ggtitle("winter_a")

cowplot::plot_grid(plt_baba_exo_distance,
                   plt_baba_exo_kuroshio_a,
                   plt_baba_exo_winter_a,
                   ncol=1)

```

## Applying a Linear Gaussian State-Space Model without exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_baba_without <- lgssm(baba_temp_ts) 
summary(res_baba_without)

# plot each of components at once
plot(res_baba_without)
```

## Applying a Linear Gaussian State-Space Model with exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_baba_with <- lgssm(temp_data=baba_temp_ts,
                          exo_data=baba_exo_ts) 
summary(res_baba_with)
# plot each of components at once
plot(res_baba_with)
```

## time series Cross-Validation

```{r message = FALSE, warning = FALSE, echo = TRUE}
# ts CV of the model without exogenous variables

## Generate list of train/test data and their indexes
# 年単位のtsCV用の訓練/テストデータの作成方法
# 1) 1998年1月から2016年12月までを訓練データ、2017年１月からの1年分をテストデータ
# 2) 1998年1月から2017年12月までを訓練データ、2018年１月からの1年分をテストデータ
# ...
# 6) 1998年1月から2017年12月までを訓練データ、2022年１月からの1年分をテストデータ
# 以上の処理をts_cv_folds()関数で自動作成
folds_6years_without <- ts_cv_folds(temp_data=baba_temp_ts,
                                 exo_data = NULL,
                                 initial = 228, # 228 monthly units from 1998-Jan to 2016-Dec 
                                 horizon = 12,
                                 step = 12,
                                 fixed_window = FALSE,
                                 allow_partial = FALSE) 

## Performe tsCV
cv_meta_without <- rolling_origin_tsCV(folds_6years_without, fold_ids=seq(1:6)) # Hold on a few minutes
#cv_meta_without <- rolling_origin_tsCV(folds_6years_without, fold_ids=1) # Hold on a few minutes

cv_without <- map(cv_meta_without, pluck, "CV", .default = NULL)

cv_without_tidy <- cv_without %>%
  map(~ as.data.frame(.x) %>%tibble::rownames_to_column(var = "set")) %>%
  purrr::list_rbind(names_to = "cv_id") %>%
  mutate(Model="Without") %>%
  dplyr::relocate(Model,cv_id, set)  # 識別列を先頭に

print(cv_without_tidy)


#----------------------------------------------------------
# ts CV of the model with exogenous variables
## Generate list of train/test data and their indexes
folds_6years_with <- ts_cv_folds(temp_data=baba_temp_ts,
                     exo_data = baba_exo_ts,
                        initial = 228, # 228 monthly units from 1998-Jan to 2016-Dec 
                        horizon = 12,
                        step = 12,
                        fixed_window = FALSE,
                        allow_partial = FALSE) 

## Performe tsCV
cv_meta_with <- rolling_origin_tsCV(folds_6years_with, fold_ids=seq(1:6)) # Hold on a few minutes

cv_with <- map(cv_meta_with, pluck, "CV", .default = NULL)
#cv_with <- base::lapply(cv_meta_with, `[[`, "CV")

cv_with_tidy <- cv_with %>%
  map(~ as.data.frame(.x) %>%tibble::rownames_to_column(var = "set")) %>%
  purrr::list_rbind(names_to = "cv_id") %>%
  mutate(Model="With") %>%
  dplyr::relocate(Model,cv_id, set)  # 識別列を先頭に

print(cv_with_tidy)

cv_tidy <- bind_rows(cv_without_tidy,cv_with_tidy)
cv_tidy$Model <- factor(cv_tidy$Model,
                        levels=c("Without","With"))

plot_MAE <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MAE)) +
  geom_boxplot()

plot_MAPE <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MAPE)) +
  geom_boxplot()

plot_MASE_naive <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MASE_naive)) +
  geom_boxplot()

plot_MASE_snaive <- ggplot(data=cv_tidy,
                   aes(x=Model,y=MASE_snaive)) +
  geom_boxplot()

cowplot::plot_grid(plot_MAE,plot_MAPE,plot_MASE_naive,plot_MASE_snaive,nrow=1)
```

## ==============================================

# Appendix: Example Data and the `sst_jma2ts()` Function

## Example Data 2: Monthly air temperature at Mt. Akadake, Hokkaido, Japant
## Duration: August 2010 to August 2023
## The data originated from: https://www.biodic.go.jp/moni1000/findings/data/index.html. 
## The ThermoSSM package provides the file "example_monthly_temp.csv", containging example data. Format of the csv file is as follows: 

```text
2010,8,13.6
2010,9,6.8
2010,10,0.2
...
```

```{r message = FALSE, warning = FALSE, echo = TRUE}
example_csv_path <- system.file("extdata", "example_monthly_temp.csv", package = "ThermoSSM")
example_temp <- readr::read_csv(example_csv_path)
head(example_temp)


# Converting tidy object to ts object
MtAkadake_ts <- ts(matrix(example_temp$Temp),
  start = c(example_temp$Year[1],example_temp$Month[1]),
  frequency = 12
)

colnames(MtAkadake_ts) <- "Temp"
head(MtAkadake_ts)

frequency(MtAkadake_ts)
start(MtAkadake_ts)
end(MtAkadake_ts)

# Perform linear Gaussian state-space modelling
res_MtAkadake <- lgssm(MtAkadake_ts)
summary(res_MtAkadake)
plot(res_MtAkadake)
```

## Example Data 3: Daily sea-surface temperature off the southern coast of Ibaraki Prefecture, Japan
### Duration: January 1982 to December 2025
### This dataset consists of publicly available daily sea-surface temperature data released by the Japan
### Meteorological Agency (JMA) (https://www.jma.go.jp/jma/indexe.html).

```{r message = FALSE, warning = FALSE, echo = TRUE}
# A zoo object of daily sea-surface temperature (SST)
# off the southern coast of Ibaraki Prefecture, Japan
data(ibaraki_sst) 
head(ibaraki_sst)

# Convert the daily zoo object to a monthly ts object
monthly_ibaraki_sst_ts <- ThermoSSM::zoo_daily2ts_monthly(ibaraki_sst)
head(monthly_ibaraki_sst_ts)

# Converting from daily zoo object to monthly ts object
monthly_ibaraki_sst_ts <- ThermoSSM::zoo_daily2ts_monthly(ibaraki_sst)
head(monthly_ibaraki_sst_ts)

# Perform linear Gaussian state-space modelling
res_ibaraki_sst <- lgssm(monthly_ibaraki_sst_ts)
summary(res_ibaraki_sst)
plot(res_ibaraki_sst)
```

## Utility Function: sst_jma2ts()

### This function downloads publicly available daily mean sea-surface temperature (SST) data
### for Japanese coastal waters provided by the Japan Meteorological Agency (JMA),
### and returns the corresponding monthly mean time series as an object of class ts.

```{r message = FALSE, warning = FALSE, echo = TRUE}
#' @param sea_area_id
#' Numeric sea area ID. The default is NULL.
#' For example, 138 corresponds to the coastal waters off southern Ibaraki, Japan.
#' A list of sea area IDs and their corresponding regions is available at:
#' \url{https://www.data.jma.go.jp/kaiyou/data/db/kaikyo/series/engan/eg_areano.html}
```

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Download SST data for the southern coastal waters of Ibaraki Prefecture, Japan
sst_138_ts <- ThermoSSM::sst_jma2ts(sea_area_id = 138) 
head(sst_138_ts)

```

