---
title: "Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"
author: "Akira Hirao"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
#    toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm
vignette: >
  %\VignetteIndexEntry{"Quick tutorial for linear Gaussian state-space analysis of temperature time series with ThermoSST"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Set Ennvironment

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(ThermoSSM)
library(forecast)

library(readr)
library(dplyr)
library(ggplot2)

```

# Practice I: Analysis of Monthly Air Temperature

## Data set
## The ThermoSSM package provides an example dataset with csv format. It contains a monthly temperature time series measured at Mount Akadake, Hokkaido, Japan. The data originated from: https://www.biodic.go.jp/moni1000/findings/data/index.html. Contents of the csv file as follows:
## Year, Month, Temp
## 2010,8,13.6
## 2010,9,6.8
## 2010,10,0.2
## ...

```{r message = FALSE, warning = FALSE, echo = TRUE}
example_csv_path <- system.file("extdata", "example_monthly_temp.csv", package = "ThermoSSM")
example_temp <- readr::read_csv(example_csv_path)
head(example_temp)


# Converting tidy object to ts object
MtAkadake_ts <- ts(matrix(example_temp$Temp),
  start = c(example_temp$Year[1],example_temp$Month[1]),
  frequency = 12
)

colnames(MtAkadake_ts) <- "Temp"
head(MtAkadake_ts)

frequency(MtAkadake_ts)
start(MtAkadake_ts)
end(MtAkadake_ts)
```
## Plotting Monthly Time Series of Air Temperature

```{r message = FALSE, warning = FALSE, echo = TRUE}
plot_temp_MtAkadake <- forecast::autoplot(MtAkadake_ts) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Monthly mean air temperature at Mt. Akadake, Hokkaido, Japan")

plot(plot_temp_MtAkadake)
# The example data includes missing observation values
```
## Plotting Time Series of Temperature Anomalies

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
example_anomaly <- ThermoSSM::monthly_anomaly(MtAkadake_ts)

monthly_example_anomaly_plot <- forecast::autoplot(example_anomaly) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Temperature anomaliese")
  
plot(monthly_example_anomaly_plot)
```

## Plotting Monthly Seasonal Cycle

```{r message = FALSE, warning = FALSE, echo = TRUE}
monthly_seasonal_cycle_MtAkadake <- ThermoSSM::mean_seasonal_cycle(MtAkadake_ts) 
summary(monthly_seasonal_cycle_MtAkadake)

plt_monthly_seasonal_cycle_MtAkadake <- ggplot(
  data = monthly_seasonal_cycle_MtAkadake,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of temperature",
    x = "Month",
    y = expression(Temperature~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  )


plot(plt_monthly_seasonal_cycle_MtAkadake)
```

## Applying a Linear Gaussian State-Space Model

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_MtAkadake <- lgssm(MtAkadake_ts) 
summary(res_MtAkadake)
```

## Plotting Level, Drift, Seasonal, and Auto-Regressive Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot each of components at once
plot(res_MtAkadake)
```

## Simple Model Diagnostics

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Normality of standardized residuals
std_obs_resid <- stats::rstandard(res_MtAkadake$kfs, type = "recursive")

# Residual diagnostics
forecast::checkresiduals(std_obs_resid)
# Ljung–Box test: if P > 0.05, there is no evidence of significant autocorrelation in the residuals
# Inspect the diagnostic plots 
# top: residual time series;
# bottom left: residual autocorrelation;
# bottom right: histogram of residuals to check for outliers or other anomalies
```

## Estimated Parameters and Components

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Parameters including process errors and observation error
params <- ThermoSSM::extract_param(res_MtAkadake)
params

# Smoothing estimates
alpha_hat <- res_MtAkadake$kfs$alphahat
head(alpha_hat)


#　Smoothing estimate of level component
level_ts <- ThermoSSM::extract_level_ts(res_MtAkadake)

#　Smoothing estimate of drift component
drift_ts <- ThermoSSM::extract_drift_ts(res_MtAkadake)

# Average drift rate per year
mean_drift_year <- mean(drift_ts) * 12
print(mean_drift_year)

# Average drift rate per year from 2015 to 2018
drift_per_year_2015_2017 <- window(drift_ts,
                               start=c(2015,1),
                               end=c(2017,12)
                               ) %>%  mean()*12
print(drift_per_year_2015_2017)

# Average drift rate per year from 2015 to 2018
drift_per_year_2020_2022 <- window(drift_ts,
                               start=c(2020,1),
                               end=c(2022,12)
                               ) %>%  mean()*12
print(drift_per_year_2020_2022)

```

## Plotting Level Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_ci <- plot(res_MtAkadake,
                     components = "level",
                     ci = TRUE,
                     ci_level = 0.95
                     )

plot(plt_level_ci)
```

## Plotting Drift Component with 95% Confidence Interval

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_drift_ci <- plot(res_MtAkadake,
                     components = "drift",
                     ci = TRUE,
                     ci_level = 0.95
                     )

plot(plt_drift_ci)
```

# Practice II: Analysis of Monthly Sea-Surface Temperature with Exogenous Variables

## Dataset is originated from Baba et al. (2024)

## Baba, S., Ishii, H., and Yoshiyama, T. (2024).  
## Estimating sea temperature trends using a linear Gaussian state-space model in Jogashima, Kanagawa, Japan.
## Bulletin of the Japanese Society of Fisheries Oceanography, 88, 190–199 (in Japanese with English abstruct)
## https://doi.org/10.34423/jsfo.88.3_190
## Supplementary code and test data are available on GitHub repository:  
## https://github.com/logics-of-blue/sea-temperature-trend-jogashima

## loading dataset and converting ts object

```{r message = FALSE, warning = FALSE, echo = TRUE}
baba_data <- readr::read_csv("https://raw.githubusercontent.com/logics-of-blue/sea-temperature-trend-jogashima/refs/heads/main/src/ssm-sample-data.csv")

# time　　　：日付(論文では0.5か月だが、本サンプルコードでは月単位データとした)
# temp　　　：水温
# distance　：黒潮流軸までの距離
# kuroshio_a：黒潮流路がA型のときに1をとるフラグ
head(baba_data)

# 秋冬の黒潮Aフラグの追加
baba_data <- 
  baba_data %>% 
  mutate(winter = as.numeric(lubridate::month(as.Date(time)) %in% c(11, 12, 1, 2, 3)),
         winter_a = winter * kuroshio_a)

# winter　：秋冬に1をとるフラグ
# winter_a：秋冬かつ、黒潮がAであるときに1をとるフラグ
head(baba_data)

# ts型に変換
# 0.5か月単位データの場合はfrequency = 24と指定する
baba_data_ts <- 
  baba_data %>% 
  dplyr::select(-time, -winter) %>%             # 日付列と秋冬フラグを削除
  ts(start = c(1998, 1), frequency = 12) # ts型に変換(1998年1月開始。12か月1周期)
head(baba_data_ts)
```

## Splitting temperature data and exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
# ts object of temperature time series
baba_temp_ts <- baba_data_ts[,"temp"]

# ts object of exogenous variables
baba_exo_ts <- baba_data_ts[,c("distance","kuroshio_a", "winter_a")]

```

## Plotting Time Series of Temperature 

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_baba_temp <- forecast::autoplot(baba_temp_ts) +
  labs(y = expression(Temperature~(degree*C)), 
       x = "Time") +
  ggtitle("Monthly mean sea-surface temperature")

plot(plt_baba_temp)
```

## Plotting Time Series of Exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_baba_exo_distance <- forecast::autoplot(baba_exo_ts[,"distance"]) +
  labs(x = "Time", y = "Distance") +
  ggtitle("Distance")

plt_baba_exo_kuroshio_a <- forecast::autoplot(baba_exo_ts[,"kuroshio_a"]) +
  labs(x = "Time", y = "Category") +
  ggtitle("kuroshio_a")

plt_baba_exo_winter_a <- forecast::autoplot(baba_exo_ts[,"winter_a"]) +
  labs(x = "Time",y = "Category") +
  ggtitle("winter_a")

cowplot::plot_grid(plt_baba_exo_distance,
                   plt_baba_exo_kuroshio_a,
                   plt_baba_exo_winter_a,
                   ncol=1)

```

## Applying a Linear Gaussian State-Space Model without exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_baba_without <- lgssm(baba_temp_ts) 
summary(res_baba_without)

# plot each of components at once
plot(res_baba_without)
```

## Applying a Linear Gaussian State-Space Model with exogenous variables

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_baba_with <- lgssm(temp_data=baba_temp_ts,
                          exo_data=baba_exo_ts) 
summary(res_baba_with)
# plot each of components at once
plot(res_baba_with)
```
## time series Cross-Validation

```{r message = FALSE, warning = FALSE, echo = TRUE}
# ts CV of the model without exogenous variables
folds_6years_without <- ts_cv_folds(temp_data=baba_temp_ts,
                                 exo_data = NULL,
                                 initial = 228, # 228 monthly units from 1998-Jan to 2016-Dec 
                                 horizon = 12,
                                 step = 12,
                                 fixed_window = FALSE,
                                 allow_partial = FALSE) 

cv_out_without <- tsCV(folds_6years_without, fold_ids=seq(1:6))
# Wait a few minits



folds_6years_with <- ts_cv_folds(temp_data=baba_temp_ts,
                     exo_data = baba_exo_ts,
                        initial = 228, # 228 monthly units from 1998-Jan to 2016-Dec 
                        horizon = 12,
                        step = 12,
                        fixed_window = FALSE,
                        allow_partial = FALSE) 

cv_out_with <- tsCV(folds_6years_with, fold_ids=seq(1:6))
# Wait a few minits


```
## =============================================================================================
# Appendix: Example Data and sst_jma2ts function

## Example Data 1: Monthly air temperature at the submit of Mt. Fuji, Japan 
### Duration: from 1937-Jul to 2025-Dec
### The dataset is publicly available temperature data released by the Japan Meteorological Agency (JMA) (https://www.jma.go.jp/jma/indexe.html).
```{r message = FALSE, warning = FALSE, echo = TRUE}
data(fuji_temp) # ts object of monthly air temperature at the summit of Mt. Fuji, Japan
head(fuji_temp)
```

## Example Data 2: Daily sea-surface temperature off the southern coast of Ibaraki Prefecture, Japan
### Duration: from 1982-Jan to 2025-Dec
### The dataset is publicly available temperature data released by the Japan Meteorological Agency (JMA) (https://www.jma.go.jp/jma/indexe.html).

```{r message = FALSE, warning = FALSE, echo = TRUE}
# zoo object of daily sea surface temperature (SST) off the southern coast of Ibaraki Prefecture, Japan
data(ibaraki_sst) 
head(ibaraki_sst)

# Converting from daily zoo object to monthly ts object
monthly_ibaraki_sst_ts <- ThermoSSM::zoo_daily2ts_monthly(ibaraki_sst)
head(monthly_ibaraki_sst_ts)
```

## Utility function of jma_sst2ts()

## This function downloads publicly available daily mean sea-surface temperature (SST) data for Japanese coastal waters
## provided by the Japan Meteorological Agency (JMA),and returns the monthly average data as a \code{ts} object.

#' @param sea_area_id
#' Numeric sea area ID. The default is NULL
#' For example, 138 corresponding to the coastal sea off southern Ibaraki, Japan.
#' A list of sea area IDs and their corresponding regions is available at:
#' \url{https://www.data.jma.go.jp/kaiyou/data/db/kaikyo/series/engan/eg_areano.html}

```{r message = FALSE, warning = FALSE, echo = TRUE}
# DL of SST off sourthern coastal waters of Ibaraki Prefecture, Japan
sst_138_ts <- ThermoSSM::sst_jma2ts(sea_area_id = 138) 
head(sst_138_ts)

```

